% Bibliography Section
\begin{thebibliography}{00}
\bibitem{krizhevsky2012imagenet} A. Krizhevsky, I. Sutskever, and G. E. Hinton, "ImageNet classification with deep convolutional neural networks," in \textit{Advances in Neural Information Processing Systems}, 2012, pp. 1097-1105.

\bibitem{simonyan2014very} K. Simonyan and A. Zisserman, "Very deep convolutional networks for large-scale image recognition," arXiv preprint arXiv:1409.1556, 2014.

\bibitem{gpiosenka2021} G. Piosenka, "Musical Instruments Image Classification," Kaggle, 2021. [Online]. Available: https://www.kaggle.com/datasets/gpiosenka/musical-instruments-image-classification

\bibitem{he2015delving} K. He, X. Zhang, S. Ren, and J. Sun, "Delving deep into rectifiers: Surpassing human-level performance on ImageNet classification," in \textit{Proceedings of the IEEE International Conference on Computer Vision}, 2015, pp. 1026-1034.

\bibitem{he2016deep} K. He, X. Zhang, S. Ren, and J. Sun, "Deep residual learning for image recognition," in \textit{Proceedings of the IEEE Conference on Computer Vision and Pattern Recognition}, 2016, pp. 770-778.

\bibitem{ioffe2015batch} S. Ioffe and C. Szegedy, "Batch normalization: Accelerating deep network training by reducing internal covariate shift," in \textit{International Conference on Machine Learning}, 2015, pp. 448-456.

\bibitem{srivastava2014dropout} N. Srivastava, G. Hinton, A. Krizhevsky, I. Sutskever, and R. Salakhutdinov, "Dropout: A simple way to prevent neural networks from overfitting," \textit{Journal of Machine Learning Research}, vol. 15, no. 1, pp. 1929-1958, 2014.

\bibitem{deng2009imagenet} J. Deng, W. Dong, R. Socher, L.-J. Li, K. Li, and L. Fei-Fei, "ImageNet: A large-scale hierarchical image database," in \textit{IEEE Conference on Computer Vision and Pattern Recognition}, 2009, pp. 248-255.

\bibitem{zeiler2014visualizing} M. D. Zeiler and R. Fergus, "Visualizing and understanding convolutional networks," in \textit{European Conference on Computer Vision}, 2014, pp. 818-833.

\bibitem{selvaraju2017grad} R. R. Selvaraju, M. Cogswell, A. Das, R. Vedantam, D. Parikh, and D. Batra, "Grad-CAM: Visual explanations from deep networks via gradient-based localization," in \textit{Proceedings of the IEEE International Conference on Computer Vision}, 2017, pp. 618-626.

\bibitem{tan2019efficientnet} M. Tan and Q. V. Le, "EfficientNet: Rethinking model scaling for convolutional neural networks," in \textit{International Conference on Machine Learning}, 2019, pp. 6105-6114.

\bibitem{zoph2018learning} B. Zoph and Q. V. Le, "Neural architecture search with reinforcement learning," in \textit{International Conference on Learning Representations}, 2017.
\end{thebibliography}
